Variance reduction for Markov chains via martingale representations

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
$$

$$
b = \gamma(2K_2^2(L+m) + d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
$$

$$
\rho = \exp\left\{\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
$$
where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K > 0$ is arbitrary
\end{lem}

@@

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from auxiliary lemma:
\begin{lem}
\label{lem:fp_covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then it holds for some constant $C > 0$
\begin{equation}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
\end{equation}
For the stationary distribution it holds
\begin{equation}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
\]
By Hoelder inequality,
\begin{eqnarray*}
\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
& \leq & 2\rho^{s/2}V^{1/2}(x)
\end{eqnarray*}
yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$.
\end{proof}

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity,
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
\end{multline*}
where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Assume also that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
\end{lem}
\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

@@

 Then for any Lipschitz function $f$, $\forall N $

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}

>>>>

\begin{lem}
\label{lem:v_ergodicity}
 Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
$$
d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{n}V(x)
$$
with $V(x) = 1 + \|x\|^2$ and constants
$$
C = \left(1+\pi_\gamma(V)\right)\left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
$$

$$
b = \gamma\left(2K_2^2(L+m) + d + m\right); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
$$

$$
\rho = \exp\left\{\gamma\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
$$
where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K > 0$ is arbitrary
\end{lem}

@@

\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}. We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain. We start from auxiliary lemma:
\begin{lem}
\label{lem:fp_covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then it holds with constant $C > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity} that
\begin{equation}
\label{eq:fp_cov}
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
\end{equation}
For the stationary distribution it holds
\begin{equation}
\label{eq:fp_cov_stationary}
|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
\end{equation}
\end{lem}
\begin{proof}
Note that
\[
|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)|^2 \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
\]
By Hoelder inequality,
\begin{eqnarray*}
\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
& \leq & 2\rho^{s/2}V^{1/2}(x)
\end{eqnarray*}
yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$.
\end{proof}

\begin{lem}
\label{lem:covariance}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$. Then
\begin{equation}
\label{eq:cov_v_ergodic}
\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x) + 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
\end{equation}
with constants $C > 0$ and $0 < \rho < 1$ from \eqref{eq:v_ergodicity}.
\end{lem}
\begin{proof}
Note that
\[
\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
\]
Due to $V-$ergodicity \eqref{eq:v_ergodicity},
\[
\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
\]
To bound the first term note that
\begin{multline*}
\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
\\
\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
\leq \{\eqref{eq:fp_cov}\}\leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\int\limits_{\rset^d} V(y) |\P^k(x,dy)-\pi(dy)| \leq 
\\
\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V} \leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
\end{multline*}
where $\left|\PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|$ can be bounded using \eqref{eq:fp_cov_stationary}. 
\end{proof}
Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
\begin{lem}
Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Let $\tilde{f}(x) = f(x) - \pi(f)$ and assume that $n(1-\rho^{1/2}) > 1$. Then
\begin{equation}
\label{eq:vanilla_var}
\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
\end{equation}
with constant $C_1 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
\end{lem}

\begin{proof}
Note that
\begin{eqnarray*}
\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
\end{eqnarray*}
Now we bound first and second sum using lemma~\ref{lem:covariance}:
\[
S_1 \leq \frac{2}{n}C\pi(V)\|\tilde{f}\|^2_{V^{1/2}} + \frac{\rho^N C^2V(x)\left(\rho^NV(x) + 2\|\tilde{f}\|^2_{V^{1/2}}\right)}{n^2(1-\rho)}
\]
\begin{eqnarray*}
S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
\end{eqnarray*}
Hence \eqref{eq:vanilla_var} follows.
\end{proof}

@@

 Then for any Lipschitz function $f$ and for any $ N \geq 1$,

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(f)\right)\right\}
